---
title: "DSO 545: Statistical Computing and Data Visualization"
author: "Amlan Limaye, Evan Aronson, Jia Luo, Su Lei, Wenjie Xu"
date: "Thursday, February 19, 2015"
output: pdf_document
---

### What will we learn?
    1. Intro to plyr package
    2. Theory and concept of plyr
    3. 12 core functions of plyr
    4. plyr Progress Bar
    5. Helper functions
    6. Intro to dplyr
    7. dplyr chain function
    8. Additional resources
    
    _________________________________________
    
Good morning class, let's begin today's lecture. 

We have here Amlan, Evan, Su, Wenjie today to introduce a R package that will make a data scientist's life, like yourselfs easier. 

The package we are going to talk about is PLYR 

**(action) write on the board**

Q Anyone heard of or used this package before???

In previous exercises a few class ago, we have all seen the use of ddply in subsetting and summarising data. 

Fundamentally, as the author "Hadley Wickham" (also) the author of ggplot2, puts it, this packages aims to accomplish the goal of "split- apply-combine" strategy often used in data analysis. 

**(action)write on the board Split-Apply-Combine**

There are similar existing tools that can do SAC such as:

 - Excel's pivot tables
 - SQL group by operator
 - "by argument" to many SAS procedures
 
Let's try it out in R and see for yourself what PLYR can do 

1. Load PLYR in library
```{r}
library(plyr)

#optional to load ggplot2 
library(ggplot2)
```

2. We will use the plyr dataset 'baseball' today. It contains around 15 years of data, and 23 variables. But we will mainly focus on

  - id: player id
  - year: year of the record
  - rbi: runs batted in 
  - ab: number of times the player faced the pitcher

```{r}
?baseball
```

3. Lets look at the performance of a batter over his career. So first we want to know the number of years the player started playing. 

For a single player, Ruthba01

**(action)double click the "baberuth" data frame to show **
```{r}
baberuth <- subset(baseball, id == "ruthba01")
baberuth <- transform(baberuth, cyear = year - min(year) + 1)
```

4. Lets graph it have a look how well Babe Ruth did. Here we have the number batted over number of times he batted. 
```{r}
baberuth$rbi_ab <- baberuth$rbi/baberuth$ab
ggplot(baberuth, aes(x=cyear, y=rbi_ab))+
  geom_line()
```

5. What should we do if we want to do this for every one of the batters?

**(action) Can anyone suggest any way to do this in R?**

Here is a teaser, of how we can manipulate our data. How about the plots? How can we run a plot for each 1228 batter? 
```{r}
## Here is how we can easily do in R
baseball_2 <- ddply(baseball, .(id), transform,
                  cyear = year - min(year) + 1)

head(baseball_2)

#number of unique batters
length(unique(baseball$id))
```

Next, Amlan will start by explaining the concept of PLYR and piece together the tools and process to solve data analysis problems similar to the batter one. 

Wenjie will introduce the 12 core functions in the PLYR package and progress bar. 

Evan will introduce the helper functions. 

And Su will wrap up with additional useful information including dplyr and chain function. 




